---
title: "USArrest_PCA_example"
author: ""
date: "10/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PCA Example on US Arrest data

This is making use of a built in example data set USArrests

```{r}
data(USArrests)
head(USArrests)

```

We we will scale or standardize the data.  In scaling, we subtract the mean and divide by the standard deviation
so the variables all have mean of zero and stddeve=1

This is done because of the wide variation in the scale of measurement

```{r}
scArrests=apply(USArrests,2,scale)
mean(scArrests)
sd(scArrests)
head(scArrests)
```
Note that the var-cov matrix of the scaled data is actually now a correlation matrix

We can now run the PCA

```{r}

z2=prcomp(USArrests,scale=TRUE)
summary(z2)
```

Let's plot the first two PCA axes scores

```{r}
require(ggplot2)

State<-rownames(USArrests)
PC1=z2$x[,1]
PC2=z2$x[,2]
PC3=z2$x[,3]
PC4=z2$x[,4]
myScores=data.frame(PC1=PC1,PC2=PC2,PC3=PC3,PC4=PC4)

#quartz()
#windows()


```



```{r}
print(z2$rotation)
```

```{r}
par(mfrow=c(2,2))
barplot(z2$rotation[,1])
barplot(z2$rotation[,2])
barplot(z2$rotation[,3])
barplot(z2$rotation[,4])


```




```{r}

barplot(z2$rotation[,1])
barplot(z2$rotation[,2])

```




```{r}

p<-ggplot(myScores, aes(PC1,PC2)) + 
  modelr::geom_ref_line(h = 0) +
  modelr::geom_ref_line(v = 0) +
   geom_text(aes(label = State), size = 3) +
  xlab("First Principal Component") + 
  ylab("Second Principal Component") 

print(p)

```



x axis: murder,assault,rape; high left/low right
y axis: urban population; high bottom/low top

Top right: low urban population, low murder,assault,rape
Bottom right: high urban population, low murder,assauly,rape
Bottom left: high urban population, high murder,assault,rape
Top left: low urban population, high murder,assault,rape








```{r}

barplot(z2$rotation[,3])
barplot(z2$rotation[,4])


```


Looking at cumulative varience explained PC1 and PC2 are the most.




```{r}
p<-ggplot(myScores, aes(PC3,PC4)) + 
  modelr::geom_ref_line(h = 0) +
  modelr::geom_ref_line(v = 0) +
   geom_text(aes(label = State), size = 3) +
  xlab("Third Principal Component") + 
  ylab("Fourth Principal Component") 

print(p)

```


x axis: rape; low left/high right
y axis: Murder,Assault; low murder;high assault;bottom/high murder;low assault;top

Top right: high rape/high murder;low assault
Bottom right: high rape/low murder;high assault
Bottom left: low rape/low murder;high assault
Top left: low rape/high murder;low assault




```{r}

pr.out=prcomp(USArrests,scale=TRUE)
pr.out$rotation


```



```{r}

biplot(pr.out, scale=0)


```


```{r}

pr.out$rotation=-pr.out$rotation
pr.out$x=-pr.out$x
plot1 <- biplot(pr.out,scale=0)
plot1



```



```{r}
# 1. Open jpeg file
jpeg("biplot_PCA_USArrests.jpg", width = 900, height = 900)
# 2. Create the plot
biplot(pr.out,scale=0)
# 3. Close the file
dev.off()

```




